---
title: "Crime Rates"
---

Follow along with the demo to learn how to calculate crime rates using R and the Tidyverse. We'll also introduce a new library, janitor, which helps make column names easier to work with. You should get started by loading the tidyverse in a new R block, then install and load janitor.

### Task 1: Load the tidyverse and install and load janitor
```{r}
library(tidyverse)
install.packages("janitor")
library(janitor)
```



### Task 2: Load the crime dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_annual_crime.csv
```{r}
crime_data <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_annual_crime.csv") |> clean_names()
```



### Task 3: Use janitor to clean up the column names



### Task 4: Calculate the total crime rate for each year (picking our unit!)

```{r}
crime_data |> mutate(total_rate = grand_total/population * 10000)
```


### Task 5: Calculate the violent and property crime rates for each year (same unit)

```{r}
crime_data |> mutate(violent_rate = violent_crime_total/population * 10000)
crime_data |> mutate(property_rate = property_crime_total/population * 10000)
```


### Task 6: Describe our results

Write a few sentences about what you found in the data. What years had the highest crime rates? What years had the lowest crime rates? Does the pattern match your expectations?

While looking at the data, I noticed that total crime, violent crime, and property crime were all down significantly in 2020 compared to 1975. Other than a couple hiccups in the early 80s and 90s, the criime has been on a steady decrease. This is contrary to what I expected, as I thought crime had always gone up.


When you are done, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.

