---
title: "MathScores"
output: html_document
---

---
title: "Identifying Schools with Biggest Differences in Math Test Scores, Part 2"
author: "YOUR NAME"
---

```{r}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
options(scipen = 999)
```

### Introduction

We'll do what we did in class, but with a twist: instead of Montgomery County, Pa., you'll be working with data from Montgomery County, Md. Same basic idea, though: linear regression looking at free and reduced-price meals and math test scores.

Just as in the exercise, you will create a scatterplot to visualize the relationship between the FARMS (Free and Reduced-price Meal System) percentage and the percentage of students who met or exceeded math expectations. You'll then use linear regression to identify the schools that show the biggest differences between their actual and expected math test scores and evaluate the results of your model.

### Loading the Data

First, let's load the data into a data frame using the read_csv() function from the tidyverse library.

```{r}
schools <- read_csv("https://raw.githubusercontent.com/dwillis/jour405_files/refs/heads/main/montgomery_md.csv")
```
### Task: Create a Scatterplot and Describe the Results (5 points)

First, let's create a scatterplot to visualize the relationship between the FARMS percentage and the percentage of students who met or exceeded math expectations. We'll use ggplot2 to create the plot and add a linear regression line.

```{r}
schools |>
  ggplot(aes(x = `FARMS Pct`, y = `% Met or Exceeded Math Expectations`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "FARMS Percentage", y = "% Met or Exceeded Math Expectations", title = "Scatterplot of Math Expectations vs FARMS Percentage")
```
Based on what you see here, what's the relationship between FARMS percentage and math test scores? Compare this scatterplot to the one you made in the class activity.

This one was very different then the one we made in class, as this scatterplot has a strong negative relationship, meaning that a lot of the schools that had a high percentage of students on free or reduced lunch had a lower average math score. In the Montgomery County, PA scatterplot, the line of best fit showed a positive relationship, albeit a very weak one.


### Task: Linear Regression Model (7 points)

Next, you'll perform a linear regression analysis using the lm() function. We'll set the percentage of students who met or exceeded math expectations as the dependent variable, and the FARMS percentage as the independent variable. In other words, if we know the FARMS percentage for a school, how reliably can we predict its math score?

```{r}
model <- lm(`% Met or Exceeded Math Expectations` ~ `FARMS Pct`, data = schools)

summary(model)
```

Provide the model's output for the three statistics mentioned in the class activity.

The first thing we look at is the p value. Considering how small it is here, it is significant. This is a switch from the other activity in which the p value was insignificant. In addition, the adjusted R-Squared value is significantly higher in this model (0.6734) compared to the in class activity which had a value of 0.0254. When we look at the coefficiant for the FARMS percentage here, we see that it is negative, meaning we can rule that it is going to be lower than the other activity that had a positive coefficiant.

Based on what you've learned so far, what does it tell us about the nature of the relationship between FARMS percentage and math scores? Is this model worth exploring further? Why or why not?

This model is definitely worth exploring further considering the significance of all three of the important statistics that we looked at. As answered above, the numbers confirmed the negative relationship that we observed in the scatterplot.

### Task: Identifying Schools with Biggest Residuals (3 points)

Let's calculate and examine the residuals next. 

```{r}
schools$predicted <- predict(model)

# Calculate the residuals
schools <- schools |> 
  mutate(residual = `% Met or Exceeded Math Expectations` - predicted)

# Sort the data frame by the absolute value of the residuals in descending order
schools <- schools |>
  mutate(abs_residual = abs(residual)) |>
  arrange(desc(abs_residual))

```

Look at the absolute values of the residuals; what are the largest differences between the predicted result and the actual math scores? Pick two schools with the largest residuals. Find the zip codes where they are located and using the CensusReporter.org website, describe the most interesting demographic data from each zip code.

The schools with the largest residuals were Wayside and Carderock Springs. They both had numbers that were higher than anticipated (positive residuals). Wayside is in Potomac in a zip code where the median income is over double what the average median income is for Maryland. The story is similar at Carderock Springs which is in Bethesda and has a similar story to Potomac where the median income is over double the Maryland median income.

## Task: Evaluate the model (5 points)

Describe what conclusions you'd draw from this model's results and how it would influence a story pitch about the school board member and math scores.

This model's results are obviously very different then the results from the Pennsylvania model. More research would need to be done before writing a story because the numbers are so different. However, if you are just writing about Maryland, you would pitch that schools with more students on Free and Reduced Lunch are struggling more on math testing. This would be a story that would interest the public and the school board because of the difference in their schools. It is also important to note that Pennsylvania does not have a county wide school district like Maryland does. This would also factor in to the story.
